---
title: "Viridis Demo"
output: html_document
---

```{r include = FALSE, warning = FALSE}
library(viridis)
```


# How It Works

The code below demonstrates two color palettes in the [viridis](https://github.com/sjmgarnier/viridis) package. Each plot displays a contour map of the Maunga Whau volcano in Auckland, New Zealand.

## Viridis colors

```{r}
image(volcano, col = viridis(200))
```

## Magma colors

```{r}
image(volcano, col = viridis(200, option = "A"))
```




# Code Chunks

The code below demonstrates the Magma palette in the [viridis](https://github.com/sjmgarnier/viridis) package. It displays a contour map of the Maunga Whau volcano in Auckland, New Zealand.

## Magma colors

```{r fig.cap = "The Maunga Whau volcano, Auckland."}
image(volcano, col = viridis(200, option = "A"))
```




# Inline Code

```{r include = FALSE}
# colorFunc <- "heat.colors"
colorFunc <- "terrain.colors"
# colorFunc <- "topo.colors"
# colorFunc <- "cm.colors"
# colorFunc <- "rainbow"
```

Base R comes with many functions for generating colors. The code below demonstrates the `r colorFunc` function. 

## `r colorFunc`

```{r fig.cap = "The Maunga Whau volcano.", echo = FALSE}
image(volcano, col = get(colorFunc)(200))
```




# Tables

Several packages support making beautiful tables with R, such as

* [xtable](https://cran.r-project.org/web/packages/xtable/)
* [stargazer](https://cran.r-project.org/web/packages/stargazer/)
* [pander](http://rapporter.github.io/pander/)
* [tables](https://cran.r-project.org/web/packages/tables/)
* [ascii](http://eusebe.github.io/ascii/)
* etc.

It is also very easy to make tables with knitr's `kable` function:

```{r echo = FALSE, results = 'asis'}
library(knitr)
kable(mtcars[1:5, ], caption = "A knitr kable.")
```




# Markdown Basics

Markdown provides an easy way to make standard types of formatted text, like

- *italics*
- **bold**
- `code`
- [links](rmarkdown.rstudio.com)
- etc.

But did you know that you can also use R Markdown's markdown to make

- Latex equations, $E = mc^{2}$
- And bibliographies 
    