---
title: "PBHL621 data cleaning practice"
author: "Ethan Walker"
date: "10 Feb 2020"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

R Markdowns are better than R scripts. They are easier to note/annotate,
you don't have to deal with working directories, and you can output or "knit" the
entire file to either HTML, PDF, or Word.

If you use Markdowns within an R Project, it is an easy way to keep your code
organized for a particular project or class.

I'll start by showing you how to set up a Project and copy my code to a Markdown.

Then we'll run through some data cleaning and summary techniques.

First, install the packages in the chunk below:

## Install and load libraries
```{r, message=FALSE}
# The "#" at the beginning of a line means the code will not run
# It's a helpful way to comment/annotate your code, or save code to run later

#install.packages("tidyverse")
#install.packages("readxl")
#install.packages("naniar")
#install.packages("lubridate")
#install.packages("gmodels")

library(tidyverse)
library(readxl)
library(naniar)
library(lubridate)
library(gmodels)
```

Next, download some cheatsheets: Help > Cheatsheets > Data Transformation & Data Visualization

Now, on to the data.

The dataset we will work with this week is a small subset of the National Survey 
of Children's Health (NSCH). I have made some intentional errors in the dataset. 
I'll walk through some example code with you to clean and present the data,
then give you time to work through similar issues on your own.

## First, load the data
### Side note: in a Markdown, you can run code in chunks, line-by-line, or by highlighting
```{r}
# Get in the habit of saving objects using "<-", not with "="
nsch_practice <- read_xlsx("Input/nsch_practice.xlsx")

# data has 100 rows (observations) and 27 columns (variables)
```


## Initial glance/summary of the data
```{r}
summary(nsch_practice)
```


Using the "summary()" function gives a first glance at the data.
Initially, we see that some of the variables seem off. Things like ID, State, 
and participant sex are being treated as numeric when they should be factors (categorical).
Participant age is being treated as character, when it should be numeric.

We can use other functions and techniques to help us look more in-depth at individual variables.
For example, let's work with the "language" variable first (HHLANGUAGE):


## Look at individual variables/columns
```{r}
table(nsch_practice$HHLANGUAGE)
```

We can see that this variable is not cleaned up yet, so we'll work on that next.


## Cleaning data and coding using "pipes"

"Piping" means using "%>%" to carry the code over to the next line. This enables 
you to run chunks of code at once with numerous functions, while also saving 
your work as a new dataset. Piping saves time and space, and is a more efficient 
way to code in R.
```{r}
# Start with the full, raw practice dataset and save it as a new dataset
# Always save as a new dataset -- never override your existing dataset
nsch_cleaning_1 <- nsch_practice %>% 
   # the "count()" function creates a table, similar to what we used above with "table()"
   count(HHLANGUAGE)
nsch_cleaning_1


# Next, let's fix some of the odd values in our variable
nsch_cleaning_1 <- nsch_practice %>% 
   # Use the "mutate()" function to change variables or create new variables
   # Here, we use the "if_else()" function within "mutate()" to fix our variable
   mutate(language = if_else(HHLANGUAGE == "??", "99", HHLANGUAGE)) %>% 
   count(language)
nsch_cleaning_1
# Now we see the "?" is fixed - now coded as 99


# Add to our code to fix the "0" answer
nsch_cleaning_1 <- nsch_practice %>% 
   mutate(language = if_else(HHLANGUAGE == "??" | HHLANGUAGE == "0", 
                             "99", HHLANGUAGE)) %>% 
   count(language)
nsch_cleaning_1
# The "??" and "0" are now coded as 99


# Next, let's recode the variable using the "fct_collapse()" function
nsch_cleaning_1 <- nsch_practice %>% 
   mutate(language = if_else(HHLANGUAGE == "??" | HHLANGUAGE == "0", 
                             "99", HHLANGUAGE),
          language = fct_collapse(language,
                                  "english" = c("1", "english"),
                                  "spanish" = c("2", "spanish"),
                                  "french" = "3",
                                  "NA" = "99")) %>% 
   count(language)
nsch_cleaning_1


# Saving the dataset - don't finish by running "count()"
# Note the dimensions of the dataset in the Environment window
nsch_cleaning_1 <- nsch_practice %>% 
   mutate(language = if_else(HHLANGUAGE == "??" | HHLANGUAGE == "0", 
                             "99", HHLANGUAGE),
          language = fct_collapse(language,
                                  "english" = c("1", "english"),
                                  "spanish" = c("2", "spanish"),
                                  "french" = "3",
                                  "NA" = "99"))
```


## Working with numeric and categorical data
```{r}
nsch_clean <- nsch_cleaning_1 %>% 
   # select specific variables
   select(HHID, SC_SEX, SC_AGE_YEARS, language) %>% 
   # give the variables new names
   # I prefer creating new variables over renaming variables
   mutate(id = as.factor(HHID),
          sex = as.factor(SC_SEX),
          age_child = as.numeric(SC_AGE_YEARS)) %>% 
   select(id, sex, age_child, language) %>% 
   ## you can also select by subtracting variables
   # Use "count()" to check the new variables and see if they needed further cleaning
   count(sex)
nsch_clean


# change specific values to NA
nsch_clean <- nsch_cleaning_1 %>% 
   select(HHID, SC_SEX, SC_AGE_YEARS, language) %>% 
   mutate(id = as.factor(HHID),
          sex = as.factor(SC_SEX),
          age_child = as.numeric(SC_AGE_YEARS)) %>% 
   select(id, sex, age_child, language) %>% 
   # use "replace_with_na" to change values to NA
   replace_with_na(sex, replace = list(sex = c(0, 7, 99))) %>% 
   replace_with_na(age_child, replace = list(age_child = c(99))) %>% 
   count(sex)
nsch_clean


# create new categorical age variables
nsch_clean <- nsch_cleaning_1 %>% 
   select(HHID, SC_SEX, SC_AGE_YEARS, language) %>% 
   mutate(id = as.factor(HHID),
          sex = as.factor(SC_SEX),
          age_child = as.numeric(SC_AGE_YEARS)) %>% 
   select(id, sex, age_child, language) %>% 
   # use "replace_with_na" to change values to NA
   replace_with_na(sex, replace = list(sex = c(0, 7, 99))) %>% 
   replace_with_na(age_child, replace = list(age_child = c(99))) %>% 
   # recode sex variable
   # change age to categorical/factor
   # collapse categories of a factor
   mutate(sex = factor(sex, levels = c("1", "2"), labels = c("male", "female")),
          age_child_cat = cut(age_child, breaks = c(0, 4, 8, 12, 16, 18),
                              labels = c("<4", "4-7", "8-11", "12-15", "16+"),
                              right = FALSE),
          age_child_cat_2 = fct_collapse(age_child_cat,
                                         "<12" = c("<4", "4-7", "8-11"),
                                         "12+" = c("12-15", "16+")))

summary(nsch_clean)
```
  
 
## Save the cleaned dataset
```{r, include=FALSE, eval=FALSE}
write_rds(nsch_clean, "Output/nsch_clean.rds")

write_csv(nsch_clean, "Output/nsch_clean.csv")
```
 

Go to your Output folder and open the CSV to check it out.
Notice anything off about the "age_child_cat" variable/values?
Let's fix that and re-save the dataset. This is a good example of thinking
through how you name variables, datasets, and even values in R. Some things
have downstream impacts that you don't initially realize.

Next, let's start with a fresh Global Environment (use the broom).
Then, load your clean dataset:
```{r}
nsch_clean <- read_rds("Output/nsch_clean.rds")
```


## Summarizing categorical data
```{r}
# Make a table and save it
nsch_table <- table(nsch_clean$sex, nsch_clean$age_child_cat,
                    dnn = c("sex", "age cat"))


nsch_table # simple counts within each cell
prop.table(nsch_table) # proportions: all cells add to 1
prop.table(nsch_table, 1) # proportions: rows add to 1
prop.table(nsch_table, 2) # proportions: columns add to 1


margin.table(nsch_table) # count: all cells
margin.table(nsch_table, 1) # count: rows 
margin.table(nsch_table, 2) # count: columns


# This functions comes from a separate package (gmodels)
# It gives more robust tables
CrossTable(nsch_clean$sex, nsch_clean$age_child_cat, prop.chisq = FALSE)


# Make a table with 3 variables
nsch_table_3 <- table(nsch_clean$sex, nsch_clean$age_child_cat, nsch_clean$language,
                    dnn = c("sex", "age", "language"))
nsch_table_3
```


## Summarizing numeric data
```{r}
nsch_summary_numeric <- nsch_clean %>% 
   group_by(sex) %>% 
   summarize("Mean Age" = mean(age_child))
   #summarize("Mean Age" = mean(age_child, na.rm = TRUE))
nsch_summary_numeric


# Do you not want the NAs in the table? Filter them out
nsch_summary_numeric <- nsch_clean %>% 
   group_by(sex) %>% 
   summarise("Mean Age" = mean(age_child, na.rm = TRUE)) %>% 
   filter(!is.na(sex))
nsch_summary_numeric


# Add more summary stats
nsch_summary_numeric <- nsch_clean %>% 
   group_by(sex) %>% 
   summarise("Mean Age" = mean(age_child, na.rm = TRUE),
             "SD" = sd(age_child, na.rm = TRUE),
             "n" = n(),
             "Min" = min(age_child, na.rm = TRUE),
             "25%" = quantile(age_child, 0.25, na.rm = TRUE),
             "Median" = median(age_child, na.rm = TRUE),
             "75%" = quantile(age_child, 0.75, na.rm = TRUE),
             "Max" = max(age_child, na.rm = TRUE)) %>% 
   filter(!is.na(sex))
nsch_summary_numeric
```


######### If we have time ##########

## Working with dates and times
### I added sample dates/times to this subset of the data
```{r}
# look at the formatting to decide how to work with the variables
head(nsch_practice$DATE, 10)
# date is in a "ymd" format
head(nsch_practice$TIME, 10)
# time had a random date added to it when it was imported

# Now we can format date and time to make them easier to work with
# Date will be the easier one to deal with
nsch_datetimes <- nsch_practice %>% 
   mutate(date = ymd(DATE))

head(nsch_datetimes$date, 10)

# To format time, we need to separate the time from the random date that was added
nsch_datetimes <- nsch_practice %>% 
   mutate(date = ymd(DATE)) %>% 
   separate(TIME, into = c("trash", "time"), sep = " ", remove = FALSE) %>% 
   unite("datetime", c("date", "time"), sep = " ", remove = FALSE) %>% 
   mutate(datetime = ymd_hms(datetime),
          id = as.factor(HHID)) %>% 
   select(id, datetime, date, time)


# Join the dates we just formatted with the clean dataset from above
nsch_dates_joined <- nsch_clean %>% 
   left_join(nsch_datetimes, by = "id")

head(nsch_dates_joined, 5)

# Use the new datetime variable to make new vars and arrange the data
datetime_practice <- nsch_dates_joined %>% 
   mutate(time_diff_days = as.numeric(lead(datetime) - datetime),
          time_diff_hours = time_diff_days*24,
          time_diff_min = time_diff_hours*60,
          day_of_week = as.factor(weekdays(datetime)),
          month_of_year = as.factor(month(datetime, label = TRUE, abbr = TRUE))) %>% 
   arrange(age_child_cat, datetime)

head(datetime_practice, 10)
```


## Save the cleaned, joined dataset
```{r, include=TRUE, eval=FALSE}
write_rds(datetime_practice, "Output/nsch_clean_dates.rds")

write_csv(datetime_practice, "Output/nsch_clean_dates.csv")
```


## Try to "Knit" your Markdown to an HTML document

You need to install more software/packages to knit to PDF/Word.

Check out the R Markdown cheatsheet and reference guide (Help > Cheatsheets)
for more info on Markdown options and ways to present and share your code.

If you use R a lot (dissertation?) or with groups of people, it can be helpful 
to back your code up to Github. There are tutorials on how to get that set up
online.

## Finally, here are some helpful resources:

https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html 

https://www.rdocumentation.org/packages/dplyr/versions/0.7.8 

https://www.statmethods.net/stats/frequencies.html 

Or, an entire book on R for Data Science:

https://r4ds.had.co.nz/ 

One of the best things about R is the online community. If you want to do something
specific in R, there is probably already a way to do it, and someone has probably
already shared it online. Google is a great companion to R, so don't spend too
much time struggling to figure things out before checking the web for a solution.

